Density-dependent selection and the limits of relative fitness

### 2018-03-09 manuscript.tex
l

Each iteration, adults produce propagules which disperse at random, independently of distance from their parents, and independently of each other. We assume that each adult from type $i$ produces $b_i$ propagules on average, so that the mean number of $i$ propagules dispersing to unoccupied territories is $m_i=b_in_iU/T$. The parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}. Random dispersal is then modeled using a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$ for the number $x_i$ of $i$ propagules dispersing to any particular unoccupied territory, where $l_i=m_i/U$ is the mean propagule density in unoccupied territories. The total propagule density will be denoted $L=\sum_i l_i$ and the mean total number of propagules by $M=\sum_i m_i$.

We assume that adults cannot be ousted by juveniles, so that recruitment to adulthood occurs exclusively in unoccupied territories. When multiple propagules land on the same unoccupied territory, the winner is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). Since the expected fraction of unoccupied territories with propagule composition $x_1,\ldots,x_G$ is $p_1(x_1)\cdots p_G(x_G)$, and type $i$ is expected to win a proportion $c_i x_i/\sum_j c_j x_j$ of these, type $i$'s expected territorial acquisition is given by
\begin{equation}
\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\cdots p_G(x_G), \label{eq:growthsumuncoupled}
\end{equation}
where the sum only includes territories with at least one propagule present. Since we do not consider drift here, we will not analyze the fluctuations about these two expectations.

@@

In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every type $l_i\gg 1$. From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of types present), and so the probability that type $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$. The numbers of territories won by each type $\Delta_+ n_1,\Delta_+ n_2,\ldots,\Delta_+ n_G$ then follow a multinomial distribution with $U$ trials and success probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots,\frac{c_G l_G}{\sum_j c_j l_j}$, respectively. Type $i$ is expected to win a proportion $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption), giving 
\begin{equation}
\Delta_+ n_i=\frac{c_i l_i}{\sum_j c_j l_j}U=\frac{c_i l_i}{\overline{c}L}U, \label{eq:lottery}
\end{equation}
where $\overline{c}=\sum_j c_j m_j/M$ is the mean competitive ability for a randomly selected propagule. Note that all unoccupied territories are filled in a single iteration of the classic lottery model, whereas our more general model Eq.~\eqref{eq:growthsumuncoupled} allows for territories to be left unoccupied and hence for arbitrary propagule densities $l_i$.

@@

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest. The second, $R_i c_i/\overline{c}$, represents competitive victories when the $i$ type is a rare invader in a high density population (i.e. it determines invasion fitness; \citealt{metz_1992}). The third term, $A_i c_i/\overline{c}$, represents competitive victories when the $i$ type is abundant. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $l_i/L$. If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of the per-propagule success probability $\Delta_+ n_i/m_i$ in simulations, in Eq. \eqref{eq:master} and its three components, and in the classic lottery model. In the simulations, $x_i$ values are sampled from a Poisson distribution with mean $l_i$ in $U$ territories, and then the victor in each territory is decided by random sampling weighted by the lottery win probabilities $c_ix_i/\sum_j c_j x_j$. Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$. Simulation points are almost invisible for the common type due to near exact agreement with Eq. \eqref{eq:master}. Dashed lines show the breakdown of the classic lottery model. Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ varies between $5\times 10^3$ and $10^6$.} 
\end{figure}

Eq.~\eqref{eq:master} takes a much simpler form if all types are competitively equivalent ($c_i=c$),
\begin{equation}
\Delta_+ n_i = \frac{l_i}{L}(1-e^{-L})U. \label{eq:masterequalc}
\end{equation}
Here $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of territories gained, and type $i$ receives a fraction $l_i/L$ of these. Total population density thus grows according to
\begin{equation}
\Delta N=(1-e^{-L})U-\sum_i d_i n_i \label{eq:Nmaster}
\end{equation}

@@

Equipped with our variable-density lottery model, we now start evaluating the validity of Eq.~\eqref{eq:canonical}. In this section we explore whether we should expect population density to vary as a result of selection \citep{prout_1980}. Since the idea that density does vary with selection is closely connected to the notion of ``$K$-selection'', we start by revisiting MacArthur's analysis of selection in crowded environments \citep{macarthur_1967}.

MacArthur considers a population with two types that have densities $n_1$ and $n_2$ subject to density-dependent growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
The environment is assumed to remain constant apart from the type densities. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources. This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

@@

Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967}, the meaning is that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}. This general idea is much broader than ``$K$-selection'' in the sense of selection for greater saturation density, and applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

@@

It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present. However, the Lotka-Volterra model still couples selection to population density \citep{smouse_1976}. Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$). Even though the initial and final densities of a sweep are the same, density is not constant over a sweep. Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C). Intuitively, for one type to exclude another with the same saturation density, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing a dip in $N$ over the sweep.

By contrast, if one type in our density-dependent lottery model has a $c$ advantage but the types are otherwise identical (so that each type has the same saturation density), the density trajectories converge on the line of constant density equal to the saturation density (Fig.~\ref{fig:LVvslottery}b). Selection then occurs purely along this line, similarly to Fig.~\ref{fig:Ksel}b. This occurs because $c$ does not directly affect $N$: it only affects the relative likelihood for each type to win a contested territory, not whether a territory is contested in the first place (this can be seen formally in Eq.~\eqref{eq:Nmaster}). In other words, once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model. While quite different from classical growth models like the Lotka-Volterra, this is all perfectly consistent with MacArthur's general argument.

The constant-$N$ behavior of $c$-selection arises from the role of $c$ as a trait determining relative competitive success in territorial contests. As such, this behavior is a result of a reproductive excesses. By contrast, previous models of selection-independent density either used unusual models of competition \citep{kimura1969natural,nei1971fertility}, or made restrictive parameter choices in the Lotka-Volterra model (Appendix C; \citealt{smouse_1976,mallet_2012}).

@@

Let us now turn to selection on $b$ and $d$ in our lottery model. Recall that $m_i=b_i n_i U/T$, and so $L=M/U=\overline{b}N/T$ where $\overline{b}$ is the population mean $b$. Thus, from Eq.~\eqref{eq:masterequalc} we have
\begin{equation}
\Delta n_i = \left(\frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i\right)n_i. \label{eq:bdensitydependence}
\end{equation}
It can be seen that the mortality $d$ is akin to the birth rate in Eq.~\eqref{eq:simplebirthdeath}, and so, while $d$ does affect density, selection on $d$ is density independent. Thus, $d$ sweeps follow the canonical relative fitness model exactly (Fig.~\ref{fig:DDS_lottery}).

At first glance, $b$ in Eq.~\eqref{eq:bdensitydependence} appears to be analogous to the $\delta$ in Eq.~\eqref{eq:simplebirthdeath} because it regulates density and is multiplied by the density-dependent term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$. This term declines from $\overline{b}$ at low density to zero at high density and as a result, selection on $b$ is density-dependent. The source of this density-depedence is that the selective advantage from having greater $b$ depends on the number of territories being contested; if almost all are occupied, then there is little advantage to having a greater $b$.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{DDS_lottery.pdf}
\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the selection coefficient $(\Delta n_j-\Delta n_i)/n_j$ experienced by an adaptive variant $j$ present at the same frequency as a wildtype $i$. Here $b_i=1$, $d_i=0.5$ and $c_i=1$. For $b$-selection we set $b_j=b_i(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$. $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as  available territories become scarce. The equilibrium density for the $i$ type is $\approx 0.4$.}
\end{figure}

>>>>

Each iteration, adults produce propagules which disperse at random, independently of distance from their parents, and independently of each other. We assume that each adult from type $i$ produces $b_i$ propagules on average, so that the mean number of $i$ propagules dispersing to unoccupied territories is $m_i=b_in_iU/T$. The parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}. Random dispersal is then modeled using a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$ for the number $x_i$ of $i$ propagules dispersing to any particular unoccupied territory, where $l_i=m_i/U$ is the mean propagule density in unoccupied territories. The total propagule density will be denoted $L=\sum_i l_i$.

We assume that adults cannot be ousted by juveniles, so that recruitment to adulthood occurs exclusively in unoccupied territories. When multiple propagules land on the same unoccupied territory, the winner is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_i c_i x_i$, where $c_i$ is a constant representing relative competitive ability (Fig. \ref{fig:lottery}). Since the expected fraction of unoccupied territories with propagule composition $x_1,\ldots,x_G$ is $p_1(x_1)\cdots p_G(x_G)$ where $G$ is the number of types present, and type $i$ is expected to win a proportion $c_i x_i/\sum_i c_i x_i$ of these, type $i$'s expected territorial acquisition is given by
\begin{equation}
\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_i c_i x_i} p_1(x_1)\cdots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}
Here the sum only includes territories with at least one propagule present. Since we do not consider drift here, we will not analyze the fluctuations around these two expectations.

@@

In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every type ($l_i\rightarrow \infty$ for all $i$). From the law of large numbers, the composition of propagules in each territory will not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$. Type $i$ is thus expected to win a proportion $c_i l_i/\sum_i c_i l_i$ of the $U$ available territories,
\begin{equation}
\Delta_+ n_i=\frac{c_i l_i}{\sum_i c_i l_i}U=\frac{c_i l_i}{\overline{c}L}U, \label{eq:lottery}
\end{equation}
where $\overline{c}=\sum_i c_i m_i/\sum_i m_i$ is the mean competitive ability for a randomly selected propagule. Note that all unoccupied territories are filled in a single iteration of the classic lottery model, whereas our more general model Eq.~\eqref{eq:growthsumuncoupled} allows for territories to be left unoccupied and hence also accommodates low propagule densities.

@@

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these propagules secure the territories without contest. The second, $R_i c_i/\overline{c}$, represents competitive victories on territories where only a single $i$ propagule lands, and at least one other propagule from a different type (this term dominates the growth of a rare invader in a high density population and determines invasion fitness). The third term, $A_i c_i/\overline{c}$, represents competitive victories in territories where two or more $i$ type propagules are present. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $l_i/L$. If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of Eq.~\eqref{eq:master}, the classic lottery model, and simulations. The vertical axis is per-propagule success rate for all propagules $\Delta_+ n_i/m_i$, and for the three separate components in Eq.~\eqref{eq:master}. Simulations are 	conducted as follows: $x_1,\ldots,x_G$ values are sampled $U$ times from Poisson distributions with respective means $l_1,\ldots,l_G$, and the victor in each territory is then decided by random sampling weighted by the lottery win probabilities $c_ix_i/\sum_j c_j x_j$. Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$. Simulation points are almost invisible for the common type due to near exact agreement with Eq.~\eqref{eq:master}. Dashed lines show the breakdown of the classic lottery model. Parameters: $U=10^5$ and $l_1/l_2=0.1$ is kept fixed while varying the total density $L$.} 
\end{figure}

In the special case that all types are competitively equivalent (identical $c_i$), Eq.~\eqref{eq:master} takes a simpler form,
\begin{equation}
\Delta_+ n_i = \frac{l_i}{L}(1-e^{-L})U. \label{eq:masterequalc}
\end{equation}
This formula can also be deduced directly from Eq.~\eqref{eq:growthsumuncoupled}: $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of such territories, and type $i$ is expected to receive a fraction $l_i/L$ of these. Total population density thus grows according to
\begin{equation}
\Delta N=(1-e^{-L})U-\sum_i d_i n_i \label{eq:Nmaster}
\end{equation}

\subsection*{Selection in the variable-density lottery}

We now outline the properties of selection in the variable-density lottery, starting with selection on competitive ability $c$. Eq.~\eqref{eq:Nmaster} shows that $c$ is not a density-regulating trait: $c$ only affects the relative likelihood for each type to win a contested territory, not whether a territory is contested in the first place. Selection between types which only differ in $c$ occurs without changing density $N$. However, $c$-selection is density-dependent, with the strength of selection (measured by the difference between types in the per-capita growth rate $\Delta n_i/n_i$) peaking at an intermediate density (Fig.~\ref{fig:DDS_lottery}). This intermediate peak occurs because most territories are claimed without contest at low density, whereas few unoccupied territories are available to be contested at high density.

Turning to selection on the density-regulating traits $b$ and $d$, we write Eq.~\eqref{eq:masterequalc} in the form
\begin{equation}
\frac{\Delta n_i}{n_i} = \frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i, \label{eq:bdensitydependence}
\end{equation}
where we have used that fact that $L=\overline{b}N/T$, and $\overline{b}$ is the population mean $b$. It is clear from Eq.~\eqref{eq:bdensitydependence} that selection on adult mortality $d$ is independent of density. On the other hand, $b$-selection is density-dependent. The advantage of having greater $b$ declines with density because fewer territories are available to to be claimed (Fig.~\ref{fig:DDS_lottery}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{DDS_lottery.pdf}
\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as difference in propagule success rate $\Delta n_j/m_j-\Delta n_i/n_i$ between an adaptive variant $j$ and a wildtype $i$ present at the same frequency. Here $b_i=1$, $d_i=0.5$ and $c_i=1$. For $b$-selection we set $b_j=b_i(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$. $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as  available territories become scarce. The equilibrium density for the $i$ type is $\approx 0.4$.}
\end{figure}

@@

We now start evaluating the limits of Eq.~\eqref{eq:canonical} in the context of the variable-density lottery model, focusing in this section on the issue of whether density varies as a result of selection \citep{prout_1980}. We showed in the previous section that $c$-selection in the variable-density lottery does not affect density. 
However, this is not the case in most previous models of density-dependent selection, which have been strongly influenced by the notion of ``$K$-selection'' and the associated questions about whether selection in crowded populations will tend to increase population density \citep{anderson_1971,roughgarden_1979,lande_2009}. To clarify the connection between ``$K$-selection'' and the variable density lottery model, we thus revisit MacArthur's analysis of selection in crowded populations \citep{macarthur_1967}.

MacArthur considers a population with two types that have densities $n_1$ and $n_2$ subject to density-dependent growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
The environment is assumed to remain constant apart from the type densities. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources. This defines nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

@@

In summary, MacArthur's conclusion that  ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967} means that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}. This general idea is much broader than ``$K$-selection'' in the sense of selection for greater saturation density, and applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

@@

It is obvious in the Lotka-Volterra model that selection can favor a superior competitor in a crowded population even if its saturation density is lower than that of the other types present. Nevertheless, the Lotka-Volterra model still couples population density to selection \citep{smouse_1976}. Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$). Even though the initial and final densities of a sweep are the same, density is not constant over a sweep (constant density is possible but only occurs for a highly restricted subset of $r$ and $\alpha$ values; further details in Appendix C; also see \citealt{mallet_2012}). Intuitively, for one type to exclude another with the same saturation density, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing a dip in $N$ over the sweep.

By contrast, density trajectories in the variable-density lottery model converge on the saturation density line (Fig.~\ref{fig:LVvslottery}b). Selection then occurs along this line, similarly to Fig.~\ref{fig:Ksel}b. In other words, once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model. This behavior occurs because $c$ determines relative competitive success in territorial contests. As such, it is a manifestation of a reproductive excess of propagules. Previous models of selection-independent density have either relied on unusual models of competition \citep{kimura1969natural,nei1971fertility}, or made restrictive Lotka-Volterra parameter choices (Appendix C; \citealt{smouse_1976,mallet_2012}).

@@

 Thus, $d$ sweeps follow the canonical relative fitness model exactly (Fig.~\ref{fig:DDS_lottery}).

 At first glance, $b$ in Eq.~\eqref{eq:bdensitydependence} appears to be analogous to the $\delta$ in Eq.~\eqref{eq:simplebirthdeath} because it regulates density and is multiplied by the density-dependent term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$.

@@
